Foundations of ecological and evolutionary change

### 2025-06-02 foundations_of_ecoevo.Rmd
Edits up to the Discussion

The interdependence of ecological and evolutionary processes has long been recognised [e.g., @Darwin1859; @Fisher1958; @Pelletier2009], but the rise of eco-evolutionary models, which incorporate both, is relatively recent following a widespread recognition that ecology and evolution can happen on similar timescales [@Govaert2019; @Yamamichi2023].
Currently, a universally recognised formal definition of eco-evolutionary change is lacking, with some authors broadly interpreting "eco-evolutionary dynamics" to allow for a separation of ecological and evolutionary timescales [@Lion2023] and others advocating for a more narrow interpretation in which no such separation is permitted [@Bassar2021].
@Bassar2021 recognise two types of eco-evolutionary models that follow from these interpretations.
The first type uses separate equations to model population change versus evolutionary change, thereby allowing for any number of ecological, evolutionary, or environmental feedbacks [e.g., @Lion2018; @Patel2018; @Lion2023].
The second type models population demographics as functions of quantitative traits, with ecological and evolutionary change following from demographic processes and trait distributions operating on the same timescale [e.g., @Barfield2011; @Simmonds2020; @Jaggi2024]. 
Both model types can be very general, but like all predictive models, they rely on simplifying assumptions for tractability [@Levins1966; @Luque2016].
These simplifying assumptions can be grounded in the Price equation to demonstrate accuracy and logical consistency when modelling evolutionary change [e.g., @Coulson2008; @Barfield2011; @Rees2016; @Lion2018].
For example, @Barfield2011 link their model back to @Price1970, which they consider to be a "universal law of evolution", to place their conclusions concerning stage structured evolution in the broader context of evolutionary theory. 
The role of fundamental equations is therefore important for unifying theory [@Luque2016], and we believe that a fundamental equation of eco-evolutionary change is needed.

@@

@@

From eqn 1, we can derive the most fundamental equations of population ecology (Box 2) and evolutionary biology (Box 1) through an appropriate interpretation of $z_{i}$.

@@

To recover the general equation for population ecology (Box 2), we define $z_{i}$ as the identity of $i$ belonging to the population.
In other words, we set $z_{i} = 1$ to simply indicate that $i$ is a member of the population.
Further, we assume that individuals do not change species by setting $\Delta z_{i} = 0$.
In this case,

@@

Since we assume that individuals are identical, we can drop the subscript $i$ such that $\beta_{i} = \beta$ and $\delta_{i} = \delta$.
Summing from 1 to $N$, we can rewrite the above,

@@

@@

Equation 8 conserves the total probability and now clarifies $\Omega$ as a summed trait value, which equals expected population growth at $t$ times mean trait value at $t + 1$.
This is consistent with the population ecology derivation from the previous section where $z_{i} = 1$ by definition, and $\Omega = N_{t+1}$ (note $\lambda = \bar{w}$).
We can therefore rewrite eqn 7,

@@

From eqn 1, which describes fundamental birth and death processes in a population, we can derive both the most fundamental model of population ecology (eqn 2; Box 2) and the fundamental equation of evolution (eqn 10; Box 1).
We might also consider $\Omega$ to be a quantity of ecosystem function, in that it is the summed effect of the traits ($z_{i}$) of individual entities ($i$) in the population.
For example, $z$ might be interpreted as biomass or photosynthetic rate.

>>>>

The interdependence of ecological and evolutionary processes has long been recognised [e.g., @Darwin1859; @Fisher1958; @Pelletier2009], but the rise of eco-evolutionary models, which incorporate both, is relatively recent following a widespread recognition that ecology and evolution can happen on similar timescales [@Govaert2019; @Yamamichi2023].
Currently, a universally recognised formal definition of eco-evolutionary change is lacking, with some theoreticians broadly interpreting "eco-evolutionary dynamics" to allow for a separation of ecological and evolutionary timescales [@Lion2023] and others advocating for a more narrow interpretation in which no such separation is permitted [@Bassar2021].
@Bassar2021 identify two types of eco-evolutionary models that follow from these interpretations.
The first type uses separate equations to model population change versus evolutionary change, thereby allowing for any number of ecological, evolutionary, or environmental feedbacks [e.g., @Lion2018; @Patel2018; @Lion2023].
The second type models population demographics as functions of quantitative traits, with ecological and evolutionary change following from demographic processes and trait distributions operating on the same timescale [e.g., @Barfield2011; @Simmonds2020; @Jaggi2024]. 
Both model types can be very general, but like all predictive models, they rely on simplifying assumptions for tractability [@Levins1966; @Luque2016].
These simplifying assumptions are often grounded in the Price equation to demonstrate accuracy and logical consistency when modelling evolutionary change [e.g., @Coulson2008; @Barfield2011; @Rees2016; @Lion2018].
For example, @Barfield2011 link their model back to @Price1970, which they consider to be a "universal law of evolution", to place their conclusions concerning stage structured evolution in the broader context of evolutionary theory. 
The role of fundamental equations is therefore important for unifying theory [@Luque2016], and we believe that a fundamental equation of eco-evolutionary change is needed.

@@

Also need to explain units of Omega, since there is confusion here.

@@

The foundation of eco-evolutionary change is an interaction between the demographic processes of birth and death ($\beta_{i} - \delta_{i} + 1$) and some characteristic of an individual ($z_{i} + \Delta z_{i}$).
The value of $\Omega$ is a summation of individual characteristics, which takes the same units as $z$.
From eqn 1, we can derive the most fundamental equations of population ecology (Box 2) and evolutionary biology (Box 1) through an appropriate interpretation of $z$.
Under many interpretations of $z$, we can also interpret $\Omega$ as a metric of ecosystem function.

@@

To recover the general equation for population ecology (Box 2), we define $z_{i}$ as the identity of $i$ belonging to the population.
In other words, we set $z_{i} = 1$ to simply indicate that $i$ is a member of the population.
In this case, $z$ is count, which takes the unit 1 [note that 'individual' is a label, not a unit; @NewellEtAl2019].
Further, we assume that individuals do not change species by setting $\Delta z_{i} = 0$.
In this case,

@@

@@

If we assume that individuals are identical, then we can drop the subscript $i$ such that $\beta_{i} = \beta$ and $\delta_{i} = \delta$.
Summing from 1 to $N$, we can rewrite the above,

@@

Equation 8 conserves the total probability and now clarifies $\Omega$ as a summed trait value, which has the same units as $z$ and equals expected population growth at $t$ times mean trait value at $t + 1$.
This is consistent with the population ecology derivation from the previous section where $z_{i} = 1$ by definition, and $\Omega = N_{t+1}$ (note $\lambda = \bar{w}$).
We can therefore rewrite eqn 7,

@@

From eqn 1, which describes fundamental birth and death processes in a population, we can derive both the most fundamental model of population ecology (eqn 2; Box 2) and the fundamental equation of evolution (eqn 10; Box 1).

Ecosystem function
====================================================

In some cases, we might also consider $\Omega$ to be the total contribution of a population to ecosystem function.
For example, $z$ might be interpreted as trait that underlies a function of interest such as carbon capture or photosynthetic rate.
When that the relevant characteristic ($z_{i}$) is additive across individuals  (e.g., the total amount of carbon capture in a population is the sum of carbon captured across individuals), $\Omega$ could be an appropriate metric of ecosystem function.
This will not always be the case.
The sum of some measurable traits such as leaf pH [@Cornelissen2011] will not be ecologically meaningful when summed across individuals.
And in other cases, the relationship between $\Omega$ and some ecosystem function of interest is likely to be subadditive.
For example, we could interpret $z_{i}$ as the frequency at which a pollinator $i$ visits plants (which might change from $t$ to $t+1$ for any number of reasons, including ontogenic or environmental, as reflected in $\Delta z_{i}$).
While $\Omega$ would then be the total frequency of pollinator visits in a population, pollinator saturation effects would likely have a logarithmic relationship between $\Omega$ and total seed set [e.g., @Mu2018].
While we have argued that ecological and evolutionary change can both be derived from the same fundamental relationship between birth, death, and individual characteristics, the same is not true for ecosystem function.
Nevertheless, Box 3 provides a toy example of four plants with different fitnesses and carbon captures.
We use this toy example to demonstrate a concrete example of our proposed framework.

### 2025-06-02 foundations_of_ecoevo.tex
Edits up to the Discussion

The interdependence of ecological and evolutionary processes has long
been recognised (e.g., \citeproc{ref-Darwin1859}{Darwin 1859};
\citeproc{ref-Fisher1958}{Fisher 1958};
\citeproc{ref-Pelletier2009}{Pelletier et al. 2009}), but the rise of
eco-evolutionary models, which incorporate both, is relatively recent
following a widespread recognition that ecology and evolution can happen
on similar timescales (\citeproc{ref-Govaert2019}{Govaert et al. 2019};
\citeproc{ref-Yamamichi2023}{Yamamichi et al. 2023}). Currently, a
universally recognised formal definition of eco-evolutionary change is
lacking, with some authors broadly interpreting ``eco-evolutionary
dynamics'' to allow for a separation of ecological and evolutionary
timescales (\citeproc{ref-Lion2023}{Lion et al. 2023}) and others
advocating for a more narrow interpretation in which no such separation
is permitted (\citeproc{ref-Bassar2021}{Bassar et al. 2021}). Bassar et
al. (\citeproc{ref-Bassar2021}{2021}) recognise two types of
eco-evolutionary models that follow from these interpretations. The
first type uses separate equations to model population change versus
evolutionary change, thereby allowing for any number of ecological,
evolutionary, or environmental feedbacks (e.g.,
\citeproc{ref-Lion2018}{Lion 2018}; \citeproc{ref-Patel2018}{Patel et
al. 2018}; \citeproc{ref-Lion2023}{Lion et al. 2023}). The second type
models population demographics as functions of quantitative traits, with
ecological and evolutionary change following from demographic processes
and trait distributions operating on the same timescale (e.g.,
\citeproc{ref-Barfield2011}{Barfield et al. 2011};
\citeproc{ref-Simmonds2020}{Simmonds et al. 2020};
\citeproc{ref-Jaggi2024}{Jaggi et al. 2024}). Both model types can be
very general, but like all predictive models, they rely on simplifying
assumptions for tractability (\citeproc{ref-Levins1966}{Levins 1966};
\citeproc{ref-Luque2016}{Luque 2017}). These simplifying assumptions can
be grounded in the Price equation to demonstrate accuracy and logical
consistency when modelling evolutionary change (e.g.,
\citeproc{ref-Coulson2008}{Coulson and Tuljapurkar 2008};
\citeproc{ref-Barfield2011}{Barfield et al. 2011};
\citeproc{ref-Rees2016}{Rees and Ellner 2016};
\citeproc{ref-Lion2018}{Lion 2018}). For example, Barfield et al.
(\citeproc{ref-Barfield2011}{2011}) link their model back to Price
(\citeproc{ref-Price1970}{1970}), which they consider to be a
``universal law of evolution'', to place their conclusions concerning
stage structured evolution in the broader context of evolutionary
theory. The role of fundamental equations is therefore important for
unifying theory (\citeproc{ref-Luque2016}{Luque 2017}), and we believe
that a fundamental equation of eco-evolutionary change is needed.

@@

From eqn 1, we can derive the most fundamental equations of population
ecology (Box 2) and evolutionary biology (Box 1) through an appropriate
interpretation of \(z_{i}\).

@@

To recover the general equation for population ecology (Box 2), we
define \(z_{i}\) as the identity of \(i\) belonging to the population.
In other words, we set \(z_{i} = 1\) to simply indicate that \(i\) is a
member of the population. Further, we assume that individuals do not
change species by setting \(\Delta z_{i} = 0\). In this case,

@@

Since we assume that individuals are identical, we can drop the
subscript \(i\) such that \(\beta_{i} = \beta\) and
\(\delta_{i} = \delta\). Summing from 1 to \(N\), we can rewrite the
above,

@@

@@

Equation 8 conserves the total probability and now clarifies \(\Omega\)
as a summed trait value, which equals expected population growth at
\(t\) times mean trait value at \(t + 1\). This is consistent with the
population ecology derivation from the previous section where
\(z_{i} = 1\) by definition, and \(\Omega = N_{t+1}\) (note
\(\lambda = \bar{w}\)). We can therefore rewrite eqn 7,

@@

From eqn 1, which describes fundamental birth and death processes in a
population, we can derive both the most fundamental model of population
ecology (eqn 2; Box 2) and the fundamental equation of evolution (eqn
10; Box 1). We might also consider \(\Omega\) to be a quantity of
ecosystem function, in that it is the summed effect of the traits
(\(z_{i}\)) of individual entities (\(i\)) in the population. For
example, \(z\) might be interpreted as biomass or photosynthetic rate.

@@

@@

@@

>>>>

The interdependence of ecological and evolutionary processes has long
been recognised (e.g., \citeproc{ref-Darwin1859}{Darwin 1859};
\citeproc{ref-Fisher1958}{Fisher 1958};
\citeproc{ref-Pelletier2009}{Pelletier et al. 2009}), but the rise of
eco-evolutionary models, which incorporate both, is relatively recent
following a widespread recognition that ecology and evolution can happen
on similar timescales (\citeproc{ref-Govaert2019}{Govaert et al. 2019};
\citeproc{ref-Yamamichi2023}{Yamamichi et al. 2023}). Currently, a
universally recognised formal definition of eco-evolutionary change is
lacking, with some theoreticians broadly interpreting ``eco-evolutionary
dynamics'' to allow for a separation of ecological and evolutionary
timescales (\citeproc{ref-Lion2023}{Lion et al. 2023}) and others
advocating for a more narrow interpretation in which no such separation
is permitted (\citeproc{ref-Bassar2021}{Bassar et al. 2021}). Bassar et
al. (\citeproc{ref-Bassar2021}{2021}) identify two types of
eco-evolutionary models that follow from these interpretations. The
first type uses separate equations to model population change versus
evolutionary change, thereby allowing for any number of ecological,
evolutionary, or environmental feedbacks (e.g.,
\citeproc{ref-Lion2018}{Lion 2018}; \citeproc{ref-Patel2018}{Patel et
al. 2018}; \citeproc{ref-Lion2023}{Lion et al. 2023}). The second type
models population demographics as functions of quantitative traits, with
ecological and evolutionary change following from demographic processes
and trait distributions operating on the same timescale (e.g.,
\citeproc{ref-Barfield2011}{Barfield et al. 2011};
\citeproc{ref-Simmonds2020}{Simmonds et al. 2020};
\citeproc{ref-Jaggi2024}{Jaggi et al. 2024}). Both model types can be
very general, but like all predictive models, they rely on simplifying
assumptions for tractability (\citeproc{ref-Levins1966}{Levins 1966};
\citeproc{ref-Luque2016}{Luque 2017}). These simplifying assumptions are
often grounded in the Price equation to demonstrate accuracy and logical
consistency when modelling evolutionary change (e.g.,
\citeproc{ref-Coulson2008}{Coulson and Tuljapurkar 2008};
\citeproc{ref-Barfield2011}{Barfield et al. 2011};
\citeproc{ref-Rees2016}{Rees and Ellner 2016};
\citeproc{ref-Lion2018}{Lion 2018}). For example, Barfield et al.
(\citeproc{ref-Barfield2011}{2011}) link their model back to Price
(\citeproc{ref-Price1970}{1970}), which they consider to be a
``universal law of evolution'', to place their conclusions concerning
stage structured evolution in the broader context of evolutionary
theory. The role of fundamental equations is therefore important for
unifying theory (\citeproc{ref-Luque2016}{Luque 2017}), and we believe
that a fundamental equation of eco-evolutionary change is needed.

@@

The foundation of eco-evolutionary change is an interaction between the
demographic processes of birth and death
(\(\beta_{i} - \delta_{i} + 1\)) and some characteristic of an
individual (\(z_{i} + \Delta z_{i}\)). The value of \(\Omega\) is a
summation of individual characteristics, which takes the same units as
\(z\). From eqn 1, we can derive the most fundamental equations of
population ecology (Box 2) and evolutionary biology (Box 1) through an
appropriate interpretation of \(z\). Under many interpretations of
\(z\), we can also interpret \(\Omega\) as a metric of ecosystem
function.

@@

To recover the general equation for population ecology (Box 2), we
define \(z_{i}\) as the identity of \(i\) belonging to the population.
In other words, we set \(z_{i} = 1\) to simply indicate that \(i\) is a
member of the population. In this case, \(z\) is count, which takes the
unit 1 {[}note that `individual' is a label, not a unit; Newell and
Tiesinga (\citeproc{ref-NewellEtAl2019}{2019}){]}. Further, we assume
that individuals do not change species by setting \(\Delta z_{i} = 0\).
In this case,

@@

@@

If we assume that individuals are identical, then we can drop the
subscript \(i\) such that \(\beta_{i} = \beta\) and
\(\delta_{i} = \delta\). Summing from 1 to \(N\), we can rewrite the
above,

@@

Equation 8 conserves the total probability and now clarifies \(\Omega\)
as a summed trait value, which has the same units as \(z\) and equals
expected population growth at \(t\) times mean trait value at \(t + 1\).
This is consistent with the population ecology derivation from the
previous section where \(z_{i} = 1\) by definition, and
\(\Omega = N_{t+1}\) (note \(\lambda = \bar{w}\)). We can therefore
rewrite eqn 7,

@@

From eqn 1, which describes fundamental birth and death processes in a
population, we can derive both the most fundamental model of population
ecology (eqn 2; Box 2) and the fundamental equation of evolution (eqn
10; Box 1).

\section{Ecosystem function}\label{ecosystem-function}

In some cases, we might also consider \(\Omega\) to be the total
contribution of a population to ecosystem function. For example, \(z\)
might be interpreted as trait that underlies a function of interest such
as carbon capture or photosynthetic rate. When that the relevant
characteristic (\(z_{i}\)) is additive across individuals (e.g., the
total amount of carbon capture in a population is the sum of carbon
captured across individuals), \(\Omega\) could be an appropriate metric
of ecosystem function. This will not always be the case. The sum of some
measurable traits such as leaf pH
(\citeproc{ref-Cornelissen2011}{Cornelissen et al. 2011}) will not be
ecologically meaningful when summed across individuals. And in other
cases, the relationship between \(\Omega\) and some ecosystem function
of interest is likely to be subadditive. For example, we could interpret
\(z_{i}\) as the frequency at which a pollinator \(i\) visits plants
(which might change from \(t\) to \(t+1\) for any number of reasons,
including ontogenic or environmental, as reflected in \(\Delta z_{i}\)).
While \(\Omega\) would then be the total frequency of pollinator visits
in a population, pollinator saturation effects would likely have a
logarithmic relationship between \(\Omega\) and total seed set (e.g.,
\citeproc{ref-Mu2018}{Mu et al. 2018}). While we have argued that
ecological and evolutionary change can both be derived from the same
fundamental relationship between birth, death, and individual
characteristics, the same is not true for ecosystem function.
Nevertheless, Box 3 provides a toy example of four plants with different
fitnesses and carbon captures. We use this toy example to demonstrate a
concrete example of our proposed framework.

@@

\bibitem[\citeproctext]{ref-Cornelissen2011}
Cornelissen, J. H. C., F. Sibma, R. S. P. V. Logtestijn, R. A. Broekman,
and K. Thompson. 2011.
\href{https://doi.org/10.1111/j.1365-2435.2010.01765.x}{Leaf pH as a
plant trait: Species-driven rather than soil-driven variation}.
Functional Ecology 25:449--455.

@@

\bibitem[\citeproctext]{ref-Mu2018}
Mu, J., Q. Wu, Y. Yang, M. Huang, and C. M. Grozinger. 2018.
\href{https://doi.org/10.1111/oik.04711}{Plant reproductive strategies
vary under low and high pollinator densities}. Oikos 127:1081--1094.

@@

\bibitem[\citeproctext]{ref-NewellEtAl2019}
Newell, D. B., and E. Tiesinga. 2019.
\href{https://doi.org/10.6028/NIST.SP.330-2019}{The international system
of units (SI)}.
